---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(fastDummies)
library(dplyr)
## using brfss example
brfss <-
  read.csv("http://static.lib.virginia.edu/statlab/materials/data/brfss_2015_sample.csv")

## data cleaning
brfss$COPD = ifelse(factor(brfss$COPD) == 'No', 0, 1) # reference is no
brfss$RACE = factor(brfss$RACE)
brfss$RACE <-
  relevel(brfss$RACE, ref = 'White') # reference is majority in data
brfss$AGE = factor(brfss$AGE)
brfss$SEX = ifelse(factor(brfss$SEX) == 'Female', 0, 1) # reference is majority in data

## dichotimize the variables
brfss = dummy_cols(brfss)
df = (brfss[, !(names(brfss) %in% c("RACE", "AGE"))])

outcome = 'COPD'
treatment = 'SMOKE'
covariates = names(brfss)[!names(brfss) %in% c(outcome, treatment)]

## cleaned data
df = (brfss[, !(names(brfss) %in% c("RACE", "AGE"))])

names(df) <- c("COPD", "smoke", "sex", "weight",
               "average_drinks", "race_white", "race_black",
               "race_hispanic", "race_other", "age_18to24",
               "age_25to34", "age_35to44", "age_45to54",
               "age_55to64", "age_over65")
```

# Joint variable importance plot

<!-- badges: start -->
<!-- badges: end -->

Joint variable importance plot (jointVIP) visualizes each variable's outcome importance via Pearson's correlation and treatment importance via standardized mean differences. Bias curves enable comparisons to support priotization.

## Installation

You can install the released version of jointVIP with:

``` r
install.packages("jointVIP")
```

## BRFSS Example

To demonstrate, we use the 2015 Behavioral Risk Factor Surveillance System (BRFSS) example to answer the causal question: Does smoking increase the risk of chronic obstructive pulmonary disease (COPD)? The data and background is inspired by [Clay Ford's work from University of Virginia Library](https://data.library.virginia.edu/getting-started-with-matching-methods/). First, the data is cleaned to only have numeric variables, i.e. dichodimize all factored variables. Treatment variable only contains 0 (control) and 1 (treatment).

With the cleaned data, you can specify details in the function `plot_jointVIP` like so:

```{r jointvip_specify, fig.height = 5, fig.width = 5, fig.align = "center"}
library(jointVIP)

treatment = 'smoke'
outcome = 'COPD'
covariates = names(df)[!names(df) %in% c(treatment, outcome)]

## select the pilot sample from random portion
set.seed(123485)
pilot_prop = 0.2
pilot_sample_num = sample(which(df %>% pull(treatment) == 0),
                          length(which(df %>% pull(treatment) == 0)) *
                            pilot_prop)

## set up pilot and analysis data
pilot_df = df[pilot_sample_num, ]
analysis_df = df[-pilot_sample_num, ]

## minimal example
brfss_vip = plot_jointVIP(pilot_df = pilot_df,
                          analysis_df = analysis_df,
                          treatment = treatment,
                          covariates = covariates,
                          outcome = outcome,
                          use_abs = T)

## not run
## obtain summary measures
# brfss_vip$measures
## show the joint variable importance plot
brfss_vip$VIP
```
