---
title: "Get started with jointVIP"
author: "Lauren D. Liao"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ref.bib  
vignette: >
  %\VignetteIndexEntry{Get started with jointVIP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all(".")
```

```{r download, eval=FALSE}
install.packages("jointVIP")
```

## The purpose of the jointVIP

The joint variable importance plot (jointVIP for short) is designed to help identify which variables to prioritize based on both treatment and outcome for adjustment. This plot should be used during the design stage of the observational study prior to the analysis phase.

Traditionally, balance tables and Love plots are used to show standardized mean differences and identify variables with high treatment imbalance. However, adjusting _only_ based on treatment imbalance is not advised since they may not be confounders. The jointVIP helps researchers prioritize variables that contribute marginal bias on both dimensions that can be used in observational study design. Additionally, bias curves based on the omitted variable bias framework are plotted for the ease of comparison.

## Set up for example

You can install the released version of jointVIP with: 

```{r setup, message=FALSE}
# install once its on CRAN!
# install.packages("jointVIP")

# devtools::install_github("ldliao/jointVIP")

# load jointVIP package
library(jointVIP)

# data to use for example
library(causaldata)

# matching method shown in example
library(MatchIt)
library(optmatch)

# weighting method shown in example
library(WeightIt)
library(optweight)
```


###  Data cleaning

The example data uses data from the **causaldata** package, specifically the `cps_mixtape` and `nsw_mixtape` datasets [@causaldata]. The experimental version of the `nsw_mixtape` is commonly known as `lalonde` and often an example that shows propensity score matching, with propensity score estimated from the `cps_mixtape` data [@dehejia1999causal; @lalonde1986evaluating]. 

Data set contains the following variables:

* `treat` denoting whether the person was selected in the National Supported Work Demonstration job-training program
* `age` age in years
* `educ` years of education
* `black` whether the race of the person was Black
* `hisp` whether the ethnicity of the person was Hispanic
* `marr` whether the person was married
* `nodegree` whether the person has degree or not
* `re74` real earnings in 1974
* `re75` real earnings in 1975
* `re78` outcome of interest: real earnings in 1978

Here simple data cleaning is performed to log-transform the earnings. To avoid errors, those who earned 0 is transformed as $log(1) = 0$. 
After transformation, the both data has new variables:  

* `log_re74` log-real earnings in 1974
* `log_re75` log-real earnings in 1975
* `log_re78` outcome of interest: log-real earnings in 1978

The `jointVIP` package uses `pilot_df` and `analysis_df` to denote datasets used. The `analysis_df` is the matching/weighting dataset of interest while `pilot_df` contains the external controls not used in the analysis. The `pilot_df` is used to help inform the outcome correlation and compare the omitted variable bias-based (OVB-based) standardized mean difference. If external data is not available, one may choose to sacrifice a portion of the analysis controls to form the `pilot_df`.

```{r data_cleaning}
# load data for estimating earnings from 1978
# treatment is the NSW program
pilot_df = cps_mixtape
analysis_df = nsw_mixtape

transform_earn <- function(data, variables){
  data = data.frame(data)
  log_variables = sapply(variables, 
                         function(s){paste0('log_',s)})
  data[,log_variables] =
    apply(data[,variables], 2, 
        function(x){ifelse(x == 0, 
                           log(x + 1), 
                           log(x))})
  return(data)
}

pilot_df <- cps_mixtape
pilot_df <- transform_earn(pilot_df, c('re74', 're75', 're78'))

analysis_df <- nsw_mixtape
analysis_df <- transform_earn(analysis_df, c('re74', 're75', 're78'))
```

## Specifying parameters for the `create_jointVIP()` function

To visualize for the jointVIP, following parameters must be specified:  

* `treatment` the treatment variable name; 0 for control and 1 for treated
* `outcome` the outcome of interest
* `covariates` covariates of interest that are common for the two `pilot_df` and `analysis_df` --- the variables all should occur prior to treatment and be potential confounders
* `pilot_df` pilot dataset consists of external controls
* `analysis_df` analysis dataset consists of both treated and controls of interest

The `new_jointVIP` is a `jointVIP` object.

```{r jointVIP_obj}
treatment = 'treat'
outcome = 'log_re78'
covariates = c(names(analysis_df)[!names(analysis_df) %in% c(treatment,
                                           outcome, "data_id",
                                           "re74", "re75",
                                           "re78")])

new_jointVIP = create_jointVIP(treatment = treatment,
                               outcome = outcome,
                               covariates = covariates,
                               pilot_df = pilot_df,
                               analysis_df = analysis_df)
```

## Diagnostics with a `jointVIP` object

The `summary()` function outputs the maximum absolute bias, the number of variables are above the desired bias tolerance (measured in absolute always), and the number of variables that can be plotted. 

```{r summary}
summary(new_jointVIP, 
        smd = "OVB-based",
        use_abs = TRUE,
        bias_tol = 0.01)
```

The `print()` function outputs the variables and its associated bias above the absolute `bias_tol` desired.

```{r print}
print(new_jointVIP,
      smd = "OVB-based",
      use_abs = TRUE,
      bias_tol = 0.01)
```

```{r plot, dpi=300, fig.asp = 0.75, fig.width = 6, out.width = "80%", fig.align = "center"}
plot(new_jointVIP)
```

By visualizing with the `jointVIP`, several important aspects stand out. First, the most important variables are the `log_re75` and `log_re74`. Traditional methods, such as the Love plot or balance table would indicate `nodegree` and `hisp` variables to be more important to adjust than `log_re74`, but these variables show low marginal bias contribution using the jointVIP.

## Matching examples

Following matching examples are performed to illustrate the utility of the jointVIP. Based on the plot shown above, the only variables that need adjustment are the `log_re75` and `log_re74`. Post-match results are plotted to help visualize for comparison.

### Optimal pair matching

As a simple example, optimal pair matching using Mahalanobis distance is used [@matchit; @hansen2007optmatch]. Based on the desired bias tolerance, only `log_re75` and `log_re74` are inputted into the formula.

```{r optmatch_example}
# 1:1 optimal matching w/o replacement
m.out <- matchit(
  treat ~ log_re75 + log_re74,
  data = analysis_df,
  method = "optimal",
  distance = "mahalanobis"
)
optmatch_df <- match.data(m.out)[, c(treatment, outcome, covariates)]
```


### Optimal weighting

Optimal weighting example is performed below, basing the weights upon `log_re75` and `log_re74` [@weightit]. Please see documentation on @zubizarreta2015stable for details.

```{r op}
# ordering for the weightit
ordered_analysis_df = analysis_df[order(analysis_df$treat, decreasing = T),]

optwt <- weightit(treat ~ log_re74 + log_re75,
                  data = ordered_analysis_df,
                  method = "optweight", estimand = "ATE",
                  tols = 0.005, include.obj = TRUE)
# summary(optwt)

optwt_df = ordered_analysis_df[row.names(optwt$covs),
                               covariates] * optwt$weights
optwt_df$treat = optwt$treat
optwt_df$log_re78 = analysis_df[row.names(optwt$covs), 'log_re78']
```

## Post-matching/weighting assessment with `post_jointVIP` object

Below are the examples showing how to plot after matching. The main function to use is the **create_post_jointVIP()** function, which takes in the original jointVIP object `new_jointVIP` in our example. The post-matched data frame need to be specified as `post_analysis_df` argument.

The functions: `summary()`, `print()`, and `plot()` all provide comparison between original and post jointVIPs. Note that the post-matched data frames contain the pair-matched individuals --- for post-weighted data, an additional processing step multiplying the weight by the original data.frame subsetted on all the covariates.

All methods yielded satisfactory results based on desired bias tolerance.

```{r post_opt, dpi=300, fig.asp = 0.75, fig.width = 6, out.width = "80%", fig.align = "center", message=FALSE}
post_optmatch_jointVIP <- create_post_jointVIP(new_jointVIP, 
                                               post_analysis_df = optmatch_df)

summary(post_optmatch_jointVIP)
print(post_optmatch_jointVIP)
plot(post_optmatch_jointVIP, plot_title = "Post-match jointVIP using optimal matching")
```


```{r post_wt, dpi=300, fig.asp = 0.75, fig.width = 6, out.width = "80%", fig.align = "center", message=FALSE}
post_optwt_jointVIP = create_post_jointVIP(new_jointVIP,
                                           post_analysis_df = optwt_df)
summary(post_optwt_jointVIP)
print(post_optwt_jointVIP)
plot(post_optwt_jointVIP, plot_title = "Post-weighting jointVIP using optimal weighting")
```

## References
